### 2025-11-01 thesis/implementation/index.tex
WWL impl

>>>>

WWL kernels are used to construct the distance index $d \in \mathcal{D}$
for the $\sigma$ function (see section~\ref{design:selection-function})
by computing schema distance through graph-based distance metrics. The
underlying WWL implementation used by \textsc{Natural} is the reference
implementation of WWL kernels from \citeauthor{WWL}, which is implemented in
Python. Thus \textsc{Natural} can't directly use the WWL implementation
as Rust and Python FFI is not directly possible. In order to buy into the
Python libraries and ecosystem (eg, optimal transport) and utilize the existing
implementation of the WWL kernels, writing a thin layer of Rust bindings around
\citeauthor{WWL}'s implementation was deemed most sensible. Using \texttt{pyo3}
calling Python code is made possible and let's \textsc{Natural} hook into
existing ecosystems where needed. Due to the design laid out in
section~\ref{section:design} the WWL implementation is only needed during
sampling time and not required at runtime as all distances are precomputed and
stored in a distance index, keeping the runtime portability characteristics of
using Rust that were discussed above.

% - Schema similarity using Weisfeiler-Leman graph kernel
% - O(n²) computational complexity for 100+ database schemas
% - Precomputation during initialization vs dynamic updates
% - Node labeling strategy: constraints vs data types hierarchy
% - Distance index caching: initialization time vs query performance

\paragraph{Rust-Python Integration Architecture}

The WWL kernel library used by \textsc{Natural} is implemented using
\texttt{pyo3} around the existing Python \texttt{wwl} library. The Rust
bindings are usable through the \texttt{wwl} crate. The \texttt{WWLKernel}
struct encapsulates the Python module reference and provides type-safe interfaces for
both categorical and continuous propagation modes through respective methods.
Using an existing graph library (\texttt{petgraph}) the complexity of the Rust
bindings could be kept minimal, but required conversion from Rust's
\texttt{petgraph} graph representation to Python's \texttt{igraph} graph
representation.

\paragraph{Schema Graph Representation}

The \texttt{natural-graphs} library implements the schema graph representation
discussed in section~\ref{design:graph-repr} by parsing SQL statements using the
\texttt{sqlparser} library and constructing an undirected \texttt{petgraph}
graph instance. Nodes represent tables and columns and edges capture foreign
key relationships and column ownership. Node labeling is used to encode schema
constraints hierarchically as outlined in section~\ref{design:graph-repr}.

\paragraph{Propagation Mechanisms}

The \texttt{wwl} crate supports both categorical and continuous Weisfeiler-Leman
propagation schemes. Categorical propagation operates on discrete node labels
through iterative label refinement, suitable for constraint-based schema
comparison. Continuous propagation utilizes node features encoded in matrices,
enabling similarity computation based on quantitative schema properties like
column cardinalities or data type frequencies \citep{WWL}. As \textsc{Natural}
projects the database schema into a graph representation with categorical node
labels, the categorical propagation scheme is used in \texttt{natural-graphs}
to determine schema distance. The \texttt{wwl} bindings allow configuration of
the kernel through the \texttt{KernelConfig} and \texttt{DistanceConfig}
structs, which provide control over the iteration count (which defaults to 3),
sinkhorn approximation settings etc.

\paragraph{Distance Computation and Caching}

The pairwise Wasserstein distance is computed between the current database
schema and each sample database during sampling phase. The distance index is
a \texttt{HashMap} of a sample database name to it's distance to the current
database \textsc{Natural} is running on. Caching of this computation is an
efficient strategy to minimise runtime inference time as the distance index is
static as long as database schemas are not mutated. \textsc{Natural} is storing
JSON-serialized distance indices in \texttt{Vector} for fast distance lookups
at runtime. This design prevents the computational cost of WWL distance
computations from affecting inference, reducing runtime schema comparison to
constant-time lookups for all previously indexed database combinations that
were known during sampling time. Given that the set of sample database and
target databases are usually fixed, the distance lookup becomes negligible in
terms of computational cost.

\paragraph{Integration with Example Selection}

The WWL kernel is integrated in the sampling phase, where distance indices are
computed and cached. Furthermore during example selection ($\sigma$) the
cached distance index is loaded from \texttt{Vector} for distance lookups.

The distance index is used for weighting the schema compatibility in the
selection algorithm (see section~\ref{design:selection-function}) to balance
structural similarity against question-level semantic relevance for improved
in-context learning effectiveness.
